Multi-level modeling with Openflexo/\FML

### 2022-03-16 1-main.tex
clarification between tool and tooling. Tooling was often replaced by tool(s) or toolset

% \newcommand{\noteSM}[1]{\todo[color=red!40, author=\textbf{Salvador}, inline, caption={}]{#1}}
% \newcommand{\noteJC}[1]{\todo[color=pink!40, author=\textbf{JC}, inline, caption={}]{#1}}
% \newcommand{\noteSylvain}[1]{\todo[color=green!40, author=\textbf{Sylvain}, inline, caption={}]{#1}}
% \newcommand{\noteAntoine}[1]{\todo[color=yellow!40, author=\textbf{Antoine}, inline, caption={}]{#1}}
% \newcommand{\noteFabien}[1]{\todo[color=orange!40, author=\textbf{Fabien}, inline, caption={}]{#1}}
% \newcommand{\noteJoel}[1]{\todo[color=blue!20, author=\textbf{Joel}, inline, caption={}]{#1}}
 \newcommand{\noteFahad}[1]{\todo[color=purple!20, author=\textbf{Fahad}, inline, caption={}]{#1}}

@@

\hypersetup{
  pdfauthor={Sylvain Guérin, Joel Champeau, Jean-Christophe Bach, Antoine
  Beugnard, Fabien Dagnat, Salvador Martinez},
  pdftitle={Multi-level modeling with Openflexo/\FML -- A contribution to the \mlpc},
  unicode=true,
  pdfkeywords={Multi-Level Modeling Challenge, Federation Modeling Language,
  Openflexo},
  %abstract
  pdfsubject={Model federation is a multi-model \emph{management} approach based on the use of virtual models and loosely coupled links. The models in a federation remain autonomous and represented in their original technological spaces whereas virtual models and links (which are not level bounded) serve as control components used to present different views to the users and  maintain synchronization. In this paper we tackle the \emph{MULTI process modeling challenge}, which consists in providing a solution to the problem of specifying and enacting processes. Solutions must fulfill a number of requirements for a process representation defined at an abstract process-definition level and at various more concrete domain-specific levels, resulting in a multi-level hierarchy of related models. We present a solution based on model federation and discuss the advantages and limitations of using this approach for multi-level modeling. Concretely, we use virtual models and more precisely the Federation Modeling Language ({FML}) that serves to describe them as the main building block in order to solve the process modeling challenge whereas the federation feature is used as a means to provide editing tooling for the resulting process language. Our solution is fully implemented with the Openflexo framework. It fulfills all the challenge requirements.
}
}

>>>>

% \newcommand{\noteSM}[1]{\todo[color=red!40, author=\textbf{Salvador}, inline, caption={}]{#1}}
 \newcommand{\noteJC}[1]{\todo[color=pink!40, author=\textbf{JC}, inline, caption={}]{#1}}
% \newcommand{\noteSylvain}[1]{\todo[color=green!40, author=\textbf{Sylvain}, inline, caption={}]{#1}}
% \newcommand{\noteAntoine}[1]{\todo[color=yellow!40, author=\textbf{Antoine}, inline, caption={}]{#1}}
% \newcommand{\noteFabien}[1]{\todo[color=orange!40, author=\textbf{Fabien}, inline, caption={}]{#1}}
% \newcommand{\noteJoel}[1]{\todo[color=blue!20, author=\textbf{Joel}, inline, caption={}]{#1}}
 \newcommand{\noteFahad}[1]{\todo[color=purple!20, author=\textbf{Fahad}, inline, caption={}]{#1}}

@@

\hypersetup{
  pdfauthor={Sylvain Guérin, Joel Champeau, Jean-Christophe Bach, Antoine
  Beugnard, Fabien Dagnat, Salvador Martinez},
  pdftitle={Multi-level modeling with Openflexo/\FML -- A contribution to the \mlpc},
  unicode=true,
  pdfkeywords={Multi-Level Modeling Challenge, Federation Modeling Language,
  Openflexo},
  %abstract
  pdfsubject={Model federation is a multi-model \emph{management} approach based on the use of virtual models and loosely coupled links. The models in a federation remain autonomous and represented in their original technological spaces whereas virtual models and links (which are not level bounded) serve as control components used to present different views to the users and maintain synchronization. In this paper we tackle the \emph{MULTI process modeling challenge}, which consists in providing a solution to the problem of specifying and enacting processes. Solutions must fulfill a number of requirements for a process representation defined at an abstract process-definition level and at various more concrete domain-specific levels, resulting in a multi-level hierarchy of related models. We present a solution based on model federation and discuss the advantages and limitations of using this approach for multi-level modeling. Concretely, we use virtual models and more precisely the Federation Modeling Language ({FML}) that serves to describe them as the main building block in order to solve the process modeling challenge whereas the federation feature is used as a means to provide editing tools for the resulting process language. Our solution fulfills all the challenge requirements and is fully implemented with the Openflexo framework.}
}

### 2022-03-16 2-abstract.tex
clarification between tool and tooling. Tooling was often replaced by tool(s) or toolset

Model federation is a multi-model \emph{management} approach based on the use of virtual models and loosely coupled links. The models in a federation remain autonomous and represented in their original technological spaces whereas virtual models and links (which are not level bounded) serve as control components used to present different views to the users and  maintain synchronization. In this paper we tackle the \emph{MULTI process modeling challenge}, which consists in providing a solution to the problem of specifying and enacting processes. Solutions must fulfill a number of requirements for a process representation defined at an abstract process-definition level and at various more concrete domain-specific levels, resulting in a multi-level hierarchy of related models. We present a solution based on model federation and discuss the advantages and limitations of using this approach for multi-level modeling. Concretely, we use virtual models and more precisely the Federation Modeling Language ({FML}) that serves to describe them as the main building block in order to solve the process modeling challenge whereas the federation aspect is used as a means to provide editing tooling for the resulting process language. Our solution is fully implemented with the Openflexo framework. It fulfills all the challenge requirements.

>>>>

Model federation is a multi-model \emph{management} approach based on the use
of virtual models and loosely coupled links. The models in a federation remain
autonomous and represented in their original technological spaces whereas
virtual models and links (which are not level bounded) serve as control
components used to present different views to the users and maintain
synchronization. In this paper we tackle the \emph{MULTI process modeling
challenge}, which consists in providing a solution to the problem of specifying
and enacting processes. Solutions must fulfill a number of requirements for a
process representation defined at an abstract process-definition level and at
various more concrete domain-specific levels, resulting in a multi-level
hierarchy of related models. We present a solution based on model federation
and discuss the advantages and limitations of using this approach for
multi-level modeling. Concretely, we use virtual models and more precisely the
Federation Modeling Language ({FML}) that serves to describe them as the main
building block in order to solve the process modeling challenge whereas the
federation feature is used as a means to provide editing tools for the
resulting process language. Our solution fulfills all the challenge
requirements and is fully implemented with the Openflexo framework.

### 2022-03-16 3-introduction.tex
clarification between tool and tooling. Tooling was often replaced by tool(s) or toolset

Model federation is a multi-model management approach based on the use of virtual models and loosely coupled links. The models in a federation remain autonomous and are represented in their original technological spaces whereas virtual models (also called conceptual models) and links that serve as control components used to present different views to the users and maintain synchronization.
Our solution is based on the model federation infrastructure. Concretely, we use Openflexo and its internal Federation Modeling Language (\FML), a language to create, link and manage virtual models, in order to solve the MULTI process challenge while the federation feature is used solely so as to provide tooling for the resulting process language. Our solution, which is fully implemented and executable, meets all the requirements.
%\noteFabien{Le paragraphe qui précède me semble redire des choses de la section 2, comment fait-on ?}

>>>>

Model federation is a multi-model management approach based on the use of
virtual models and loosely coupled links. The models in a federation remain
autonomous and are represented in their original technological spaces whereas
virtual models (also called conceptual models) and links that serve as control
components used to present different views to the users and maintain
synchronization. Our solution is based on the model federation infrastructure.
Concretely, we use Openflexo and its internal Federation Modeling Language
(\FML), a language to create, link and manage virtual models, in order to solve
the MULTI process challenge while the federation feature is used solely so as
to provide tools for the resulting process language. Our solution, which is
fully implemented and executable, meets all the requirements.
%\noteFabien{Le paragraphe qui précède me semble redire des choses de la
%section 2, comment fait-on ?}

### 2022-03-16 4-technology.tex
clarification between tool and tooling. Tooling was often replaced by tool(s) or toolset

This framework relies on the architecture of Fig.~\ref{fig:mf}. A federation
gathers a set of conceptual models, named \emph{virtual models} and a
set of \emph{federated models}. Each federated model pertains to a
\emph{technological space} and uses the language of its specific
paradigm while a virtual model is built using the Federation Modeling
Language (\FML). Each federated model is an autonomous
element that may evolve with its own tooling {\color{red} tooling is used in place of tool throughout the article.  I suggest clarifying it and then editing it all along the article}. The virtual models
serve as control elements binding the federated models together.
In this paper, we mainly use conceptual models and more precisely \FML, its
domain specific modeling language. The only use of the federation feature is on the tooling made for the process language.

>>>>

This framework relies on the architecture of Fig.~\ref{fig:mf}. A federation
gathers a set of conceptual models, named \emph{virtual models} and a set of
\emph{federated models}. Each federated model pertains to a \emph{technological
space} and uses the language of its specific paradigm while a virtual model is
built using the Federation Modeling Language (\FML). Each federated model is an
autonomous element that may evolve with its own tools. The virtual models serve
as control elements binding the federated models together.  In this paper, we
mainly use conceptual models and more precisely \FML, its domain specific
modeling language. The only use of the federation feature is on the toolset for
the process language.

### 2022-03-16 6-modelpresentation.tex
clarification between tool and tooling. Tooling was often replaced by tool(s) or toolset

Note that in all following figures, blue is used to represent \FML virtual
models and concepts (\textsf{VirtualModel} and \textsf{FlexoConcept} in
Fig.~\ref{fig:mm}), see
Fig.~\ref{fig:BPMNSubsetExample},~\ref{fig:MultilevelArchitecture},~\ref{fig:LinguisticAndOntologicInstantiation},~\ref{fig:BaseMetamodel},~\ref{fig:AcmeFullArchitecture},~\ref{fig:ToolingArchitecture}, while brown is used for instances of
\textsf{FlexoConcept}, see Fig.~\ref{fig:BPMNSubsetExample},~\ref{fig:MultilevelArchitecture},~\ref{fig:LinguisticAndOntologicInstantiation},~\ref{fig:AcmeFullArchitecture},~\ref{fig:ToolingArchitecture}.
% 3, 4, 5, 6, 7, 8, 9
% Expliquer que le use case XSure ne nécessitait pas l'extension de metamodèle requise pour le usecase Acme Software Developement Process

@@

\subsection{Openflexo tooling}
\label{subsec:tooling}

\begin{figure*}
 \centering
     \includegraphics[width=0.9 \textwidth]{Figures/ToolingArchitecture-1.5.pdf}
     \caption{Tooling architecture}
    \label{fig:ToolingArchitecture}
\end{figure*}

@@

We took advantage of model federation and the availability of diagraming features though the \textit{Diagraming Technology Adapter} to implement two interactive graphical tools built on top of the conceptual levels detailed in previous sections. The Fig.~\ref{fig:ToolingArchitecture} shows the overview of the tooling architecture and is detailed in the next section. A first tool offers a graphical edition of a Process type. The second one offers an enactment feature (the instantiation of a process from its process type), the ability within an enacted process to assign tasks to actors, and a graphical visualization of this process execution.

@@

The left of the figure presents the \textit{ProcessType graphical editor}.
\texttt{ProcessTypeEditor} is modeled as a virtual model declaring two model
slots (represented with bold circles). The first model slot references an
instance of Acme metamodel, while the second model slot references an instance
of diagram, conforms to the \textit{ProcessType diagram
specification}\footnote{a diagram \enquote{metamodel} which defines and specify
structure and graphical representations for edited items}. When executed, this
tool manages a graphical view for an instance of Acme metamodel (the
\textit{Acme model}) and a specific \texttt{ProcessType} instance. This tool
allows representing and editing a \texttt{ProcessType}. The highly reflective
nature of \FML and its tooling should be noted. When the drag and drop
interactor is applied for a new item, the tool allows choosing the concept type
to be instantiated. For example, when a \texttt{TaskType} is created, the user
must choose a concept inheriting from \texttt{TaskType} (in Acme use case it
can be \texttt{SETaskTask} or \texttt{CodingTaskType} or default value
\texttt{TaskType}).

@@

A demonstrative video for the developed tooling is available on our website\footnote{\url{https://research.openflexo.org/MLMChallenge.html}}, as well as download and installation instructions.

>>>>

Note that in all following figures, blue is used to represent \FML virtual
models and concepts (\textsf{VirtualModel} and \textsf{FlexoConcept} in
Fig.~\ref{fig:mm}), see
Fig.~\ref{fig:BPMNSubsetExample},~\ref{fig:MultilevelArchitecture},~\ref{fig:LinguisticAndOntologicInstantiation},~\ref{fig:BaseMetamodel},~\ref{fig:AcmeFullArchitecture},~\ref{fig:ToolingArchitecture},
while brown is used for instances of \textsf{FlexoConcept}, see
Fig.~\ref{fig:BPMNSubsetExample},~\ref{fig:MultilevelArchitecture},~\ref{fig:LinguisticAndOntologicInstantiation},~\ref{fig:AcmeFullArchitecture},~\ref{fig:ToolingArchitecture}.
% 3, 4, 5, 6, 7, 8, 9 Expliquer que le use case XSure ne nécessitait pas
% l'extension de metamodèle requise pour le usecase Acme Software Developement
% Process

@@

\subsection{Openflexo toolset}
\label{subsec:tooling}

\begin{figure*}
 \centering
     \includegraphics[width=0.9 \textwidth]{Figures/ToolingArchitecture-1.5.pdf}
     \caption{Tool architecture}
    \label{fig:ToolingArchitecture}
\end{figure*}

@@

We took advantage of model federation and the availability of diagraming
features though the \textit{Diagraming Technology Adapter} to implement two
interactive graphical tools built on top of the conceptual levels detailed in
previous sections. The Fig.~\ref{fig:ToolingArchitecture} shows the overview of
the tool architecture and is detailed in the next section. A first tool offers
a graphical edition of a Process type. The second one offers an enactment
feature (the instantiation of a process from its process type), the ability
within an enacted process to assign tasks to actors, and a graphical
visualization of this process execution.

@@

The left of the figure presents the \textit{ProcessType graphical editor}.
\texttt{ProcessTypeEditor} is modeled as a virtual model declaring two model
slots (represented with bold circles). The first model slot references an
instance of Acme metamodel, while the second model slot references an instance
of diagram, conforms to the \textit{ProcessType diagram
specification}\footnote{a diagram \enquote{metamodel} which defines and specify
structure and graphical representations for edited items}. When executed, this
tool manages a graphical view for an instance of Acme metamodel (the
\textit{Acme model}) and a specific \texttt{ProcessType} instance. This tool
allows representing and editing a \texttt{ProcessType}. The highly reflective
nature of \FML and its tools should be noted. When the drag and drop
interactor is applied for a new item, the tool allows choosing the concept type
to be instantiated. For example, when a \texttt{TaskType} is created, the user
must choose a concept inheriting from \texttt{TaskType} (in Acme use case it
can be \texttt{SETaskTask} or \texttt{CodingTaskType} or default value
\texttt{TaskType}).

@@

A demonstrative video for the developed tool is available on our
website\footnote{\url{https://research.openflexo.org/MLMChallenge.html}}, as
well as download and installation instructions.

### 2022-03-16 7-requirementssatisfaction.tex
clarification between tool and tooling. Tooling was often replaced by tool(s) or toolset

In the previous Sect.~\ref{sec:model}, we demonstrated that our solution satisfies all the challenge requirements. In this section, we summarize the different techniques used to satisfy them, without repeating all the justifications. We classify these techniques into three categories:
\begin{itemize}
    \item \emph{syntactic conformance}: the requirements are structurally satisfied when the model conforms to its metamodel,
    \item \emph{constraints checking}: the requirements are fulfilled when all instances satisfy both syntactic conformance and the evaluation of invariants expressed for related concepts,
    \item \emph{tooling}: the requirements are enforced by the implementation of the associated tooling.
\end{itemize}

@@

    \item \emph{Behavioral modeling} (\textit{B.Mod.}): a requirement is satisfied by the definition of one or more behaviors, which may be combined with constraints checking and associated tooling (for example \textbf{P17}).
\end{itemize}

>>>>

In the previous Sect.~\ref{sec:model}, we demonstrated that our solution
satisfies all the challenge requirements. In this section, we summarize the
different techniques used to satisfy them, without repeating all the
justifications. We classify these techniques into three categories:
\begin{itemize}
    \item \emph{syntactic conformance}: the requirements are structurally
      satisfied when the model conforms to its metamodel,
    \item \emph{constraints checking}: the requirements are fulfilled when all
      instances satisfy both syntactic conformance and the evaluation of
      invariants expressed for related concepts,
    \item \emph{tooling}: the requirements are enforced by the implementation
      of the associated tool.
\end{itemize}

@@

    \item \emph{Behavioral modeling} (\textit{B.Mod.}): a requirement is
  satisfied by the definition of one or more behaviors, which may be combined
  with constraints checking and associated tools (for example \textbf{P17}).
  \end{itemize}

### 2022-03-16 8-discussion.tex
clarification between tool and tooling. Tooling was often replaced by tool(s) or toolset

%\todo[inline]{on pourrait aussi répondre ici aux 3 questions posées
%explicitement (copiées dans l'intro pour le moment), a priori non, le chef
%« trouve que c'est chiant »}

%\noteJC{pour le moment, seulement réponse aux différents points, on ne voit pas
%de discussion sur les choix, sur les potentiels compromis/déficiencies}

@@

%  \noteJC{pas clair pour moi ce qui est entendu par « basic modeling constructs », ce qui est dans la partie 2 ?}

@@

  As for cross-level relationships, cross-level constraints do not have a
  specific nature. They are like any constraint a user can define by adding a
  behavior to a model element. Therefore, if a constraint is mandatory when
  establishing a relationship between elements from different levels, the user
  has to create it manually. This can be a limit of our approach: the cost of
  the flexibility of our tooling is a limited number of automated behaviors.

@@

We use the Openflexo infrastructure to build models and metamodels together,
and to provide dedicated tooling to edit a Process type, to enact a process and
to execute it. Being able to provide a solution and its associated tools
quickly and easily is a noticeable feature.  We could also have taken advantage
of the model federation to connect our solution to other tools dedicated to
process edition and to process execution (\eg a \BPMN engine). However we
already had all the necessary components to provide the aforementioned
graphical tools.

@@

Our tooling includes mechanisms to verify the models. First, syntactical
consistency is realized by cardinality checks and by typing. Second, the
constraints the users have written are continuously checked during model
edition and enactment. Thus, a part of the verification relies on the fact that
users add behaviors when they build the metamodels and the models. For
instance, \textbf{S6} requirement is constrained by an invariant implemented in
\texttt{CodeArtifact}. It ensures that \texttt{artifactType} has the right type
and that it implements the \texttt{languages} relation.
%\noteJC{updated}

>>>>

@@

@@

  As for cross-level relationships, cross-level constraints do not have a
  specific nature. They are like any constraint a user can define by adding a
  behavior to a model element. Therefore, if a constraint is mandatory when
  establishing a relationship between elements from different levels, the user
  has to create it manually. This can be a limit of our approach: the cost of
  the flexibility of our toolset is a limited number of automated behaviors.

@@

We use the Openflexo infrastructure to build models and metamodels together,
and to provide dedicated tools to edit a Process type, to enact a process and
to execute it. Being able to provide a solution and its associated tools
quickly and easily is a noticeable feature.  We could also have taken advantage
of the model federation to connect our solution to other tools dedicated to
process edition and to process execution (\eg a \BPMN engine). However we
already had all the necessary components to provide the aforementioned
graphical tools.

@@

Our toolset includes mechanisms to verify the models. First, syntactical
consistency is realized by cardinality checks and by typing. Second, the
constraints the users have written are continuously checked during model
edition and enactment. Thus, a part of the verification relies on the fact that
users add behaviors when they build the metamodels and the models. For
instance, \textbf{S6} requirement is constrained by an invariant implemented in
\texttt{CodeArtifact}. It ensures that \texttt{artifactType} has the right type
and that it implements the \texttt{languages} relation.
%\noteJC{updated}
